Open-source landscape for 3D CSEM modelling

### 2020-05-05 manuscript/Manuscript.tex
updates to introduction from RR, let's see if my comment and edit style is convenient

@@

Controlled-source electromagnetic (CSEM) measurements are a frequently applied
method in various geophysical exploration fields, such as geothermal and ground
water, oil and gas, mining, civil engineering, or geo-hazards. Modelling these
electromagnetic fields is therefore of great interest in order to understand
the measured data. Publications regarding 3D modelling in electromagnetic
methods started to appear as early as the 1970's and 1980's. These early
publications where integral equation (IE) methods, having an anomaly embedded
within a layered medium, mostly for loop-loop type transient EM measurements
\citep{GJI.74.Raiche, GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman} and
magnetotelluric (MT) measurements \citep{GEO.84.Wannamaker}.

In the 1990's computer became sufficiently powerful that 3D modelling gained
traction, which resulted amongst other in the publication of the book
\emph{Three-Dimensional Electromagnetics} by the SEG
\citep{B.SEG.99.Oristaglio}. Often cited publications from that time are
\cite{RSC.94.Mackie}, 3D MT computation; \cite{RS.94.Druskin}, frequency- and
time-domain modelling using a Yee grid and a global Krylov subspace
approximation; and \cite{RS.96.Alumbaugh, GJI.97.Newman}, low-to-high frequency
computation on massively parallel computers.

The continuous improvement of computing power and the CSEM boom in the early
2000's in the hydrocarbon industry led to a wealth of publications. The amount
of available numerical solutions can be overwhelming and is part of the reason
why there are hundreds of publications about the topic. There are the different
methods to solve Maxwell's equation, such as the IE method
\citep{GJI.74.Raiche, RS.02.Hursan, GEO.06.Zhdanov, GP.10.Tehrani,
CAG.16.Kruglyakov, MGS.17.Kruglyakov} and different variations of the
differential equation (DE) method, for instance finite differences (FD)
\citep{IEEE.66.Yee, GEO.93.Wang, RSC.94.Mackie, RS.94.Druskin, GEO.09.Streich,
CAG.13.Sommer}, and finite elementes (FE) \citep{GJI.11.Schwarzbach,
GEO.04.Commer, GEO.12.daSilva, GJI.13.Puzyrev, GJI.13.Grayver, SEG.16.Zhang},
finite volume (FV) \citep{EM.90.Madsen, ECP.07.Haber, GEO.14.Jahandari,
PIER.01.Clemens, GP.06.Mulder}. And these are just the most common ones.

There are also many different types of discretisation, where the most common
ones are regular grids (Cartesian, rectilinear), mostly using a Yee grid
\citep{IEEE.66.Yee} or a Lebedev grid \citep{CMMP.64.Lebedev}, but also
unstructured tetrahedral grids \citep{SEG.16.Zhang, CAG.17.Cai}, OcTree meshes
\citep{ECP.07.Haber}, or hexagonal meshes \citep{CAG.14.Cai}.

The biggest variety of all exists probably in the available solvers to solve
the system of linear equations; direct solvers \citep{GEO.09.Streich,
GEO.15.Grayver, GP.14.Chung, GEO.14.Jaysaval, SEG.15.Oh, GJI.18.Wang}, indirect
solvers \citep{GP.06.Mulder, GJI.15.Jaysaval}, or a combination of both,
so-called hybrid solvers \citep{GEO.18.Liu}; the solvers often use
preconditioners such as the multigrid method \citep{SIAM.02.Aruliah,
GP.06.Mulder, GJI.16.Jaysaval}.

A very well written overview up to the year 2005 of the different approaches to
3D EM modelling is given by \cite{SG.05.Avdeev}, and also by
\cite{SG.10.Borner}. In the last 15 years the publications with regards to 3D
EM  modelling grew tremendously, driven by the ever increasing computing power.
Avdeev finished his review with the following statement: «\emph{The most
important challenge that faces the EM community today is to convince software
developers to put their 3-D EM forward and inverse solutions into the public
domain, at least after some time. This would have a strong impact on the whole
subject and the developers would benefit from feedback regarding the real needs
of the end-users.}»

It is this topic that we address, open-source 3D CSEM codes. Some major
exploration service companies as well as certain research consortia and
research groups have their 3D CSEM codes. But how about codes that are
open-source? Research code was traditionally often available upon request by
the author. Codes distributed that way often come with the request to not share
the code, and they often come without a license attached at all (which means
that they are not open-source). Also, at times there can be significant hurdles
to install a code. However, the meaning of open-source evolved rapidly in the
last decades. Today, open-source generally not only means that a code is
available (with a proper license). The meaning includes much more these days,
such as that the code is hosted online, is under version control, has the
possibility to file issues and make pull requests to fix bugs. Well maintained
codes often have continuous integration that includes testing of the code base
and online hosted documentation which often includes code documentation,
examples, and tutorials. As such the term has shifted from purely open-source
code to development in the open and increasingly building a global community.
There are a number of projects within the realm of geophysical exploration,
besides the ones we present here, \mycom[DW]{Others coming to your mind?}e.g.
pyGIMLi \citep{CAG.17.Rucker} and Fatiando \citep{JOSS.18.Uieda}. Newer
\mycom[DW]{wrong term, think about it}programming environments  have also
brought much simplifications in the installation process and have most
importantly simplified it for code developers to make their codes easily
available and installable. All our presented codes are in the Python ecosystem,
and each code can be installed with a single command (which includes download
and installation).

A second topic we want to address is verification. New codes are generally
verified against analytical solutions. However, analytical and semi-analytical
solutions only exist for very simple models. There is no possibility to check
how good a computed response is, expect for comparing the result from different
modellers. If different modellers using different discretisation and
implementations of Maxwell's equations yield the same result it can give
confidence in the accuracy. Providing these two models and results from four
different codes should give a benchmark with which new codes can be compared to
and validated with.

It is worth mentioning that there are many more 3D EM codes developed in other
fields than geophysics such as communication systems (antennas, radar,
satellites), medical imaging, or civil engineering. There are various
open-source codes in these fields too, e.g., Elmer \mycom[DW]{Add 2-3 proper
references} (\href{http://www.csc.fi/elmer}{csc.fi/elmer}, couldn't find a
publication, just a poster). While these codes could potentially be used for
the same goals as presented here we restrict our review to codes purpose-built,
and ready without further adjustments, for geophysical applications.

In the following section we introduce the codes under consideration. The
numerical result consider two cases. First a layered background with vertical
transverse isotropy containing three blocks, followed by a realistic, complex
3D model.

@@

%
\tabl[btp]{comp-layered}{Comparison of used CPU and RAM and of number of cells
  used to discretize the layered model. \imycom[DW]{I will fill this table out
  once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 &  0 & 0.0 & 0 \\
  \emg3d  & 0.0 &  0 & 0.0 & 0 \\
  \petgem & 0.0 &  0 & 0.0 & 0 \\
  \simpeg & 0.0 &  0 & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%

@@

%
\tabl[btp]{comp-block}{Comparison of used CPU and RAM and of number of cells
  used to discretize the block model. \imycom[DW]{I will fill this table out
  once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 & 24 & 0.0 & 0 \\
  \emg3d  & 0.0 &  1 & 0.0 & 0 \\
  \petgem & 0.0 & 48 & 0.0 & 0 \\
  \simpeg & 0.0 &    & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%

@@

%
\tabl[btp]{comp-marlim}{Comparison of used CPU and RAM and of number of cells
  used to discretize the Marlim R3D model. \imycom[DW]{I will fill this table
  out once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 &  0 & 0.0 & 0 \\
  \emg3d  & 0.0 &  0 & 0.0 & 0 \\
  \petgem & 0.0 &  0 & 0.0 & 0 \\
  \simpeg & 0.0 &  0 & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%

>>>>

\mycom[RR]{GMD looks very interesting, my only concern is that it might not be in the scope of EM modelers, but we can probably reach modeleres from other communities as well.}

@@

Controlled-source electromagnetic (CSEM) measurements are a frequently applied method in various geophysical exploration fields, such as geothermal and groundwater, oil and gas, mining, civil engineering, or geo-hazards.
Modelling these electromagnetic fields is therefore of great interest to design survey layouts, understand the measured data and inversion purposes.
Publications regarding 3D modelling in electromagnetic methods started to appear as early as the 1970's and 1980's.
These early publications were integral equation (IE) methods, having an anomaly embedded within a layered medium, mostly for loop-loop type transient EM measurements \citep{GJI.74.Raiche, GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman} and magnetotelluric (MT) measurements \citep{GEO.84.Wannamaker}.

In the 1990's, computer became sufficiently powerful that 3D modelling gained traction, which resulted amongst other publications in the book
\emph{Three-Dimensional Electromagnetics} by the SEG
\citep{B.SEG.99.Oristaglio}.
Often cited publications from that time are \cite{RSC.94.Mackie}, 3D MT computation; \cite{RS.94.Druskin}, frequency- and time-domain modelling using a Yee grid and a global Krylov subspace approximation; and \cite{RS.96.Alumbaugh, GJI.97.Newman}, low-to-high frequency computation on massively parallel computers.

The continuous improvement of computing power and the CSEM boom in the early 2000's in the hydrocarbon industry led to a wealth of developed numerical solutions and according publications.
The most common applied methods to solve Maxwell's equation are the IE method \citep{GJI.74.Raiche, RS.02.Hursan, GEO.06.Zhdanov, GP.10.Tehrani,
CAG.16.Kruglyakov, MGS.17.Kruglyakov} and different variations of the
differential equation (DE) method, such as finite differences (FD)
\citep{IEEE.66.Yee, GEO.93.Wang, RSC.94.Mackie, RS.94.Druskin, GEO.09.Streich, CAG.13.Sommer}, finite elements (FE) \citep{GJI.11.Schwarzbach, GEO.04.Commer, GEO.12.daSilva, GJI.13.Puzyrev, GJI.13.Grayver, SEG.16.Zhang}, and finite volumes (FV) \citep{EM.90.Madsen, ECP.07.Haber, GEO.14.Jahandari, PIER.01.Clemens, GP.06.Mulder}.
There are also many different types of discretisation, where the most common ones are regular grids (Cartesian, rectilinear), mostly using a Yee grid \citep{IEEE.66.Yee} or a Lebedev grid \citep{CMMP.64.Lebedev}, but also unstructured tetrahedral grids \citep{SEG.16.Zhang, CAG.17.Cai}, hexagonal meshes \citep{CAG.14.Cai}, or OcTree meshes\citep{ECP.07.Haber}.
Very well written overviews about the different approaches to 3D EM modelling are given by \cite{SG.05.Avdeev} and \cite{SG.10.Borner}.

In the last 15 years, the publications with regards to 3D EM modelling grew tremendously, driven by the ever increasing computing power.
One reason why there are so many reports about this topic results from the variety of solution techniques for the systems of linear equations.
They can be distinguished between direct solvers \citep{GEO.09.Streich,
GEO.15.Grayver, GP.14.Chung, GEO.14.Jaysaval, SEG.15.Oh, GJI.18.Wang}, indirect solvers \citep{GP.06.Mulder, GJI.15.Jaysaval}, or a combination of both, so-called hybrid solvers \citep{GEO.18.Liu}.
The solvers often use preconditioners such as the multigrid method \citep{SIAM.02.Aruliah, GP.06.Mulder, GJI.16.Jaysaval}.

Probably the main reason for the numerous amount of code developments was already outlined by \cite{SG.05.Avdeev}, who finished his review with the following statement: «\emph{The most important challenge that faces the EM community today is to convince software developers to put their 3-D EM forward and inverse solutions into the public domain, at least after some time. This would have a strong impact on the whole subject and the developers would benefit from feedback regarding the real needs of the end-users.}»
Some major exploration service companies as well as certain research consortia and research groups own their 3D CSEM codes, but how about codes that are open-source?
Academic codes were traditionally often available upon request by authors.
Codes distributed that way were often provided with the request to not share the code, and they often come without a license attached which means that they are not open-source.
Also, at times there can be significant hurdles to install a code.

The meaning of the term open-source evolved rapidly in the last two decades.
Today, open-source generally not only means that a code is available (with a proper license), but also that the code is hosted online, is under version control, has the possibility to file issues and make pull requests to fix bugs. These are important requirements for reproducible research «\emph{to provide the means by which the reader can verify the validity of the results and make use of them in further research}» \citep{broggini2017reproducible}.
Well maintained codes often have continuous integration that includes testing of the code base and online hosted documentation, often including examples and tutorials.
As such, the term has shifted from purely open-source codes to developments in the open with increasingly building a global community.
Related projects within the realm of geophysical exploration are e.g., pyGIMLi \citep{CAG.17.Rucker} or Fatiando \citep{JOSS.18.Uieda}. \mycom[DW]{Others coming to your mind?}\mycom[RR]{no thoughts about geophysics, but there are many projects for related topics such as gmt, gis, mesh-gen}

We continue addressing the topic of open-source developments by introducing and comparing four recent projects with focus on 3D CSEM modeling.
All presented codes are in the Python ecosystem and use either the finite-difference method on structured grids or the finite-element method on unstructured tetrahedral meshes.
As newer package-management systems such as Conda, Docker, or pip significantly help developers to make their codes easily available and installable, each one can be downloaded and installed with a single command.

Furthermore, we deal with the topic of verification.
New codes are generally verified against analytical solutions.
However, analytical and semi-analytical solutions only exist for very simple models.
There is no possibility to check how good a computed response is, expect for comparing results from different modellers.
If different discretisations and implementations of Maxwell's equations yield the same result, it gives confidence in their accuracy.
We simulate EM fields for a layered background model with vertical transverse isotropy, containing three blocks, and a realistic, complex marine 3D model.
These two model designs as well as the corresponding results from four different codes provide a benchmark for new codes to be compared to and validated with.

It is worth mentioning that there are many more 3D EM codes developed in other fields than geophysics such as communication systems (antennas, radar, satellites), medical imaging, or civil engineering.
There are various open-source codes in these fields too, e.g., Elmer \imycom[DW]{Add 2-3 proper references} (\href{http://www.csc.fi/elmer}{csc.fi/elmer}, couldn't find a publication, just a poster).
While these codes could potentially be used for the same goals as presented here we restrict our review to codes purpose-built and ready without further adjustments for geophysical applications.

\imycom[RR]{I thought a lot about this but I just think the previous paragraph is not required in our introduction and haven't find a location where it fits considering different logical structures. I think it is also not important for this paper to tell "that there are more codes in EM than only geophysics", that's quite obvious. Maybe delete or shorten to 1 sentences to add Elmer and place this sentence somewhere else? I think, an appropriate list of all non-geophysical open-source EM codes should be very long anyway, we just don't know them or at least I'm not familiar with them.}

In the following section we introduce the codes under consideration.
Afterwards, we present the modeling result of the two model cases.
Finally, we discuss, conclude.\mycom[RR]{structure of the paper can be written if we are completeley clear about this and the rest of the introduction.}

@@

%
\tabl[btp]{comp-layered}{Comparison of used CPU and RAM and of number of cells
  used to discretize the layered model. \imycom[DW]{I will fill this table out
  once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{c c c c c}  %{lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 &  0 & 0.0 & 0 \\
  \emg3d  & 0.0 &  0 & 0.0 & 0 \\
  \petgem & 0.0 &  0 & 0.0 & 0 \\
  \simpeg & 0.0 &  0 & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%

@@

%
\tabl[btp]{comp-block}{Comparison of used CPU and RAM and of number of cells
  used to discretize the block model. \imycom[DW]{I will fill this table out
  once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{c c c c c}  %lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 & 24 & 0.0 & 0 \\
  \emg3d  & 0.0 &  1 & 0.0 & 0 \\
  \petgem & 0.0 & 48 & 0.0 & 0 \\
  \simpeg & 0.0 &    & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%

@@

%
\tabl[btp]{comp-marlim}{Comparison of used CPU and RAM and of number of cells
  used to discretize the Marlim R3D model. \imycom[DW]{I will fill this table
  out once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{c c c c c}  %{ls[table-format=4.2]rs[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 &  0 & 0.0 & 0 \\
  \emg3d  & 0.0 &  0 & 0.0 & 0 \\
  \petgem & 0.0 &  0 & 0.0 & 0 \\
  \simpeg & 0.0 &  0 & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%
